Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-25 ms/main.tex
revised results and discussion

% Abstract
\begin{abstract}
We study the cultural evolution of cooperation under vertical, horizontal, and oblique transmission.
Conditions are found for fixation and coexistence of cooperation and defection. 
We find that the evolution of cooperation is facilitated by horizontal transmission, especially when there is an association between cooperation and transmission, and that the effect of oblique transmission depends on the bias in horizontal transmission. Stable coexistence of cooperation and defection can occur. 
A spatial model is constructed and compared to results from an unstructured model.
Comparisons are drawn with Hamilton's rule and the concepts of relatedness and assortment.
% population mean fitness?
% TODO what about oblique?
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can reduce an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, including primates~\citep{jaeggi2013natural},  rats~\citep{rice1962altruism}, birds~\citep{stacey1990cooperative,krams2008experimental}, and lizards~\citep{sinervo2006self}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology~\citep[Appendix]{Haldane1932book}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Models
\section*{Models}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Results
\section*{Results}

In the following sections, we determine the equilibria of the model, namely, solutions of $\hat{p}' = \hat{p}$, and analyze their local stability.
We then analyze the evolution of a modifier of social association.
Finally, we demonstrate the application of the conditions derived in the previous sections to predict the outcomes of stochastic simulations in a structured population.

\subsection*{Equilibria and Stability}

To determine the equilibria in our model, we analyze the fixed points of \autoref{eq:nextgen_juveniles}.
We define $f(\hat{p}) = \bar{w}(\hat{p}' - \hat{p})$.
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics,  this simplifies to
\begin{equation} \label{eq:general_case_polynomial}
  f(\hat{p}) = \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p},
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
\beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) , \\
\beta_2 &= -\beta_1 -\beta_3 ,  \\
\beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) .
\end{aligned}
\end{equation}

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$. 
Thus, $f(\hat{p})$ becomes a quadratic polynomial,
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixA}), where
$f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big]$, so that
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
%Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{corollary:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

In the general case where $T_A \neq T_B$, the coefficient $\beta_1$ is not necessarily zero, and $f(\hat{p})$ is a cubic polynomial.
Therefore, three equilibria may exist, two of which are
$\hat{p} = 0 $ and $\hat{p} = 1$, and the third is
\begin{equation} \label{eq:general_equilibrium}
  \hat{p}^* =  
  \frac{\beta_3}{\beta_1}.
\end{equation}

Note that the sign of the  cubic (Eq.\ \ref{eq:general_case_polynomial}) at positive (negative) infinity is equal (opposite) to the sign of $\beta_1$. 
If $T_A>T_B$, then 
\begin{equation} \label{eq:beta1}
   \beta_1 < [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) 
   = (1-\alpha v)(c-b)(T_A-T_B) < 0 ,
 \end{equation}
since $c<b$ and $\alpha v < 1$. Hence the signs of the cubic at positive and negative infinity are negative and positive, respectively.
First, if $\beta_3<\beta_1$ then 
$1<\hat{p}^*$ and therefore $f'(0)<0$ and $f'(1)>0$; that is, fixation of the defector phenotype $B$ is the only locally stable legitimate (i.e. between 0 and 1) equilibrium.
Second, if $\beta_1<\beta_3<0$ then 
$0<\hat{p}^*<1$ and therefore $f'(0)<0$ and $f'(1)<0$ so that both fixations are locally stable and $\hat{p}^*$ separates the domains of attraction.
Third, if $0<\beta_3$ then 
$\hat{p}^*<0$ and therefore $f'(0)>0$ and $f'(1)<0$; that is, fixation of the cooperator phenotype $A$ is the only locally stable legitimate equilibrium.

Similarly, if $T_A<T_B$, then
\begin{equation} \label{eq:beta1_rev}
   \beta_1 > [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) 
   = (1-\alpha v)(c-b)(T_A-T_B) > 0,
 \end{equation}
since $c<b$ and $\alpha v < 1$, and the signs of the cubic at positive and negative infinity are positive and negative, respectively. 
First, if $\beta_3<0$ then $\hat{p}^*<0$ and therefore $f'(0)<0$ and $f'(1)>0$; that is, fixation of the defector phenotype $A$ is the only locally stable legitimate equilibrium.
Second, if $0<\beta_3<\beta_1$ then $0<\hat{p}^*<1$ and therefore $f'(0)>0$ and $f'(1)>0$; that is, both fixations are locally unstable and $\hat{p}^*$ is a stable polymorphic equilibrium.
Third, if $\beta_1<\beta_3$ then $\hat{p}^*>1$ and therefore $f'(0)>0$ and $f'(1)<0$, and fixation of the cooperator phenotype $A$ is the only locally stable legitimate equilibrium.

We define the \emph{cost boundaries}, $\hat\gamma_1$ and $\hat\gamma_2$,
\begin{equation} \label{eq:cost_boundaries}
\begin{aligned}
\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
\end{aligned}
\end{equation}
and the \emph{vertical transmission threshold}, $\hat v$,
\begin{equation} \label{eq:v_threshold}
\hat v = \frac{T_B - T_A}{1-T_A} \,.
\end{equation}

First, assume $T_A<T_B$.
$\beta_3<0$ requires $\hat\gamma_1<c$,
and for $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
Thus, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c$ or $v < \hat v$ and $c < \hat\gamma_2$,
and for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\hat\gamma_2 < c < \hat\gamma_1$, or $v < \hat v$ and $c < \min(\hat\gamma_1, \hat\gamma_2)$. For $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\hat\gamma_2$ or $v < \hat v$ and $\hat\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\hat\gamma_2$.

Second, when $T_A>T_B$
$\beta_3>0$ requires $\hat\gamma_1 > c$. 
For $\beta_1<\beta_3$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] < bv\alpha T_B + (1+b)(T_A-T_B)$.
Thus for $\beta_1<\beta_3$ we need $v > \hat v$ and $c < \hat\gamma_2 $ or $v < \hat v$ and $c > \hat\gamma_2$.
But $\hat{v}<0$ when $T_A > T_B$, and therefore we have $\beta_1<\beta_3$ if $c < \hat\gamma_2$. Similarly, we have $\beta_3<\beta_1$ if $c > \hat{\gamma_2}$.

@@

\begin{result}[Equilibria and stability] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the cost boundaries $\hat\gamma_1$ and $\hat\gamma_2$ and the vertical transmission threshold $\hat v$ (Eqs.~\ref{eq:cost_boundaries}-\ref{eq:v_threshold}) :

@@

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T_A \ge T_B$ and $c < \hat\gamma_1$; or 
	if \emph{(ii)} $T_A < T_B$ and $v>\hat v$ and $c < \hat\gamma_2$.
\item \emph{Fixation of defection}: 
    if \emph{(iii)} $T_A \ge T_B$ and $\hat\gamma_2 < c$; or 
	if \emph{(iv)} $T_A < T_B$ and $\hat\gamma_1 < c$.
\item \emph{Stable coexistence}: 
    if \emph{(v)} $T_A < T_B$ and $v<\hat{v}$ and $c < \hat\gamma_1$; or 
    if \emph{(vi)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_2 < c < \hat\gamma_1$.
\item \emph{Unstable coexistence}:
    if \emph{(vii)} $T_A > T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$.
\end{enumerate}

@@

Much of the literature on evolution of cooperation focuses on conditions for  initially rare cooperative phenotype to invade a population of defectors.
The next remarks deal with such conditions. The first introduces a condition on the cost of cooperation, $c$, and the second on the social association, $\alpha$.
\\

\begin{remark}[Condition for cooperation to increase from rarity: cost of cooperation]
\label{corollary:rarity}
If the initial frequency of cooperation is very close to zero, $\hat{p}_0 \approx 0$, then its frequency will increase if 
\begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
c < \hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
\end{equation} 
\end{remark}

This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection. In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias $T_A-T_B$.

Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation, $T_A$, increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), facilitating the evolution of cooperation.
However, increasing the horizontal transmission of defection, $T_B$, can increase or decrease the cost boundary, but it increases the cost boundary when it is already above one ($c<1<\hat\gamma_1$): $\partial\gamma_1/\partial T_B$ is positive when $T_A > \frac{1}{1+\alpha b v}$, which gives $\hat\gamma_1>1/v$. 
Therefore, increasing $T_B$ decreases the cost boundary and limits the evolution of cooperation, but only if $T_A < \frac{1}{1+\alpha b v}$.

Increasing the vertical transmission $v$ can either increase or decrease the boundary, depending on the horizontal transmission bias, $T_A-T_B$, because $\text{sign}(\partial \hat\gamma_1 / \partial v) = -\text{sign}(T_A-T_B)$.
When $T_A<T_B$ we get $\partial \hat\gamma_1 / \partial v >0$ and therefore, as the vertical transmission increases the cost boundary increases, making it easier for cooperation to evolve from rarity. 
In contrast, when $T_A > T_B$ we get $\partial \hat\gamma_1 / \partial v <0$ and therefore, as the vertical transmission increases the cost boundary decreases making it harder for cooperation to evolve from rarity.

When there is no horizontal transmission bias, $T_A=T_B$, the next remark applies.
\\

\begin{remark}[Symmetric horizontal transmission] 
\label{corollary:equal_transmission}
If $T=T_A=T_B$, then cooperation will take over the population from any initial frequency if
\begin{equation}
\label{eq:equal_transmission}
c < b v \cdot \frac{\alpha T}{1-T} \;.
\end{equation}
\end{remark}
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:unequal_transmission_from_rarity_general_case} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $v \alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition for $v=1$. Note that the right-hand side equals $\hat\gamma_1$ when $T=T_A=T_B$.

%In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
%A more general case can be stated for $0 < v \leq 1$.
In the next remarks we examine the effect of social association on the evolution of cooperation.
\\

\begin{remark}[No social association]
\label{corollary:alpha0}
Without social association ($\alpha=0$), cooperation can increase from rarity, and go to fixation, if there is horizontal transmission bias for cooperation, $T_A>T_B$, and
\begin{equation}
\label{eq:vert_hori_alpha0}
c < \frac{T_A - T_B}{v(1-T_B)} \;.
\end{equation}
\end{remark}

\autoref{fig:result2}a illustrates these conditions, which are obtained by setting $\alpha=0$ in Eqs.\ \ref{eq:unequal_transmission_from_rarity_general_case} and \ref{eq:cost_boundaries}.
Importantly, the benefit of cooperation, $b$, does not affect the evolution of cooperation in the absence of social association. The outcome is determined only by  cultural transmission.
\\

\begin{remark}[Perfect social association]
\label{corollary:alpha1}
With perfect social association ($\alpha=1$), cooperation will increase from rarity if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b v T_A + (T_A - T_B)}{v(1-T_B)} \;.
\end{equation}
\end{remark}
This condition is obtained by setting $\alpha=1$ in Eqs.\ \ref{eq:unequal_transmission_from_rarity_general_case} and \ref{eq:cost_boundaries}.

@@

Given the previous two remarks for no association and perfect association ($\alpha=0$ and $\alpha=1$), it is interesting to examine the general effect of social association on the evolution of cooperation.
Consider the \emph{social association boundaries}, $\hat{a}_1$ and $\hat{a}_2$,
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Then the following applies.
\\

\begin{remark}[Condition for cooperation to increase from rarity: social association]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if $\hat a_2 < \alpha$.
\end{remark}
\autoref{fig:result3}cd illustrates this condition.
\autoref{fig:result2}c demonstrates these conditions.
With horizontal transmission bias for cooperation ($T_A>T_B$), cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
With horizontal bias favoring defection ($T_A<T_B$), cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias ($T_A=T_B$) fixation of cooperation occurs if social association is high enough, $\frac{c}{b} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

Interestingly, the effect of vertical transmission on this condition depends on the horizontal transmission bias, $\partial \hat a_2 / \partial v = b T_A (T_A-T_B)$. 
That is, with horizontal bias for cooperation, $T_A>T_B$, the boundary increases with the rate of vertical transmission, impeding the evolution of cooperation.
In contrast, with horizontal bias for defection, $T_A<T_B$, the boundary decreases with the rate of vertical transmission, facilitating the evolution of cooperation.

>>>>

% Abstract
\begin{abstract}
We study the cultural evolution of cooperation under vertical and non-vertical cultural transmission.
Conditions are found for fixation and coexistence of cooperation and defection. 
The evolution of cooperation is facilitated by horizontal transmission and by an association between social interactions and horizontal transmission.
The effect of oblique transmission depends on the horizontal transmission bias.
A stable coexistence between cooperation and defection can occur.
When it does, selection favors the evolution of higher association between social interactions and horizontal transmission, which leads to an increased frequency of cooperation and higher population mean fitness.
The derived conditions are compared to outcomes of stochastic simulations of structured populations.
Comparisons are drawn with Hamilton's rule and the concepts of relatedness and assortment.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can reduce an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, including primates~\citep{jaeggi2013natural},  rats~\citep{rice1962altruism}, birds~\citep{stacey1990cooperative,krams2008experimental}, and lizards~\citep{sinervo2006self}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology~\citep[Appendix]{Haldane1932book}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Models
\section*{Models}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Results
\section*{Results}

In the following sections, we determine the equilibria of the model in \autoref{eq:nextgen_juveniles}, namely, solutions of $\hat{p}' = \hat{p}$, and analyze their local stability.
We then analyze the evolution of a modifier of social association.
Finally, we compare derived conditions to outcomes of stochastic simulations with a structured population.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolution of cooperation}

To determine the equilibria in our model, we analyze the fixed points of \autoref{eq:nextgen_juveniles}.
We define $f(\hat{p}) = \bar{w}(\hat{p}' - \hat{p})$.
Then, using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, this simplifies to
\begin{equation} \label{eq:general_case_polynomial}
\begin{aligned}
  f(\hat{p}) &= \bar{w}(\hat{p}'-\hat{p}) =
  \beta_1 \hat{p}^3 + \beta_2 \hat{p}^2 + \beta_3 \hat{p} \;,
\end{aligned}
\end{equation}
where 
\begin{equation} \label{eq:polynomial_coefficients}
\begin{aligned}
  \beta_1 &= \big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B) \;, \\
  \beta_2 &= -\beta_1 -\beta_3 \;, \\
  \beta_3 &= \alpha bvT_A - cv(1-T_B) + (T_A-T_B) \;.
\end{aligned}
\end{equation}

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$. 
Thus, $f(\hat{p})$ becomes a quadratic polynomial,
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big] \;.
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (see Appendix~\autoref{sec:appendixA}), where
$f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big]$, so that
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T) \;, \\
	f'(1) &=	-\alpha bvT + cv(1-T) \;.
\end{aligned}
\end{equation}
%Therefore with symmetric horizontal transmission ($T_A=T_B$), fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary \ref{remark:equal_transmission}, namely inequality~\ref{eq:equal_transmission}.

In the general case where $T_A \neq T_B$, the coefficient $\beta_1$ is not necessarily zero, and $f(\hat{p})$ is a cubic polynomial.
Therefore, three equilibria may exist, two of which are
$\hat{p} = 0 $ and $\hat{p} = 1$, and the third is
\begin{equation} \label{eq:general_equilibrium}
  \hat{p}^* =  
  \frac{\beta_3}{\beta_1} =
  \frac{\alpha bvT_A - cv(1-T_B) + (T_A-T_B)}{\big[c(1-v) - b (1-\alpha v)\big] (T_A-T_B)} \;.
\end{equation}

Note that the sign of the  cubic (Eq.\ \ref{eq:general_case_polynomial}) at positive (negative) infinity is equal (opposite) to the sign of $\beta_1$. 
If $T_A>T_B$, then 
\begin{equation} \label{eq:beta1}
   \beta_1 < [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) 
   = (1-\alpha v)(c-b)(T_A-T_B) < 0 \;,
 \end{equation}
since $c<b$ and $\alpha v < 1$. Hence the signs of the cubic at positive and negative infinity are negative and positive, respectively.
First, if $\beta_3<\beta_1$ then 
$1<\hat{p}^*$ and therefore $f'(0)<0$ and $f'(1)>0$; that is, fixation of the defector phenotype $B$ is the only locally stable legitimate (i.e. between 0 and 1) equilibrium.
Second, if $\beta_1<\beta_3<0$ then 
$0<\hat{p}^*<1$ and therefore $f'(0)<0$ and $f'(1)<0$ so that both fixations are locally stable and $\hat{p}^*$ separates the domains of attraction.
Third, if $0<\beta_3$ then 
$\hat{p}^*<0$ and therefore $f'(0)>0$ and $f'(1)<0$; that is, fixation of the cooperator phenotype $A$ is the only locally stable legitimate equilibrium.

Similarly, if $T_A<T_B$, then
\begin{equation} \label{eq:beta1_rev}
   \beta_1 > [c(1-\alpha v) - b(1-\alpha v)] (T_A-T_B) 
   = (1-\alpha v)(c-b)(T_A-T_B) > 0 \;,
 \end{equation}
since $c<b$ and $\alpha v < 1$, and the signs of the cubic at positive and negative infinity are positive and negative, respectively. 
First, if $\beta_3<0$ then $\hat{p}^*<0$ and therefore $f'(0)<0$ and $f'(1)>0$; that is, fixation of the defector phenotype $A$ is the only locally stable legitimate equilibrium.
Second, if $0<\beta_3<\beta_1$ then $0<\hat{p}^*<1$ and therefore $f'(0)>0$ and $f'(1)>0$; that is, both fixations are locally unstable and $\hat{p}^*$ is a stable polymorphic equilibrium.
Third, if $\beta_1<\beta_3$ then $\hat{p}^*>1$ and therefore $f'(0)>0$ and $f'(1)<0$, and fixation of the cooperator phenotype $A$ is the only locally stable legitimate equilibrium.

To present this analysis in terms of conditions on the cost of cooperation $c$, we define the cost boundaries, $\gamma_1$ and $\gamma_2$, and the vertical transmission threshold, $\hat v$,

\begin{equation} \label{eq:cost_boundaries_v_threshold} 
\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)}, \quad
\hat v = \frac{T_B - T_A}{1-T_A} \;.
\end{equation}

First, assume $T_A<T_B$.
$\beta_3<0$ requires $\gamma_1<c$.
For $\beta_3<\beta_1$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] > bv\alpha T_B + (1+b)(T_A-T_B)$.
Note that the expression in the square brackets is positive if and only if $v > \hat v$.
Thus, for $\beta_3<\beta_1$ we need $v > \hat v$ and $\gamma_2 < c$ or $v < \hat v$ and $c < \gamma_2$,
and for $0<\beta_3<\beta_1$ we need $v > \hat v$ and $\gamma_2 < c < \gamma_1$, or $v < \hat v$ and $c < \min(\gamma_1, \gamma_2)$. For $\beta_1<\beta_3$ we need $v > \hat v$ and $c<\gamma_2$ or $v < \hat v$ and $\gamma_2<c$.
However, some of these conditions cannot be met, since $v < \hat v$ implies $c<1<\gamma_2$.

Second, assume $T_A>T_B$.
$\beta_3>0$ requires $\gamma_1 > c$. 
For $\beta_1<\beta_3$ we need $c\big[v(1-T_B) + (1-v)(T_A-T_B)\big] < bv\alpha T_B + (1+b)(T_A-T_B)$.
Thus for $\beta_1<\beta_3$ we need $v > \hat v$ and $c < \gamma_2 $ or $v < \hat v$ and $c > \gamma_2$.
But $\hat{v}<0$ when $T_A > T_B$, and therefore we have $\beta_1<\beta_3$ if $c < \gamma_2$. Similarly, we have $\beta_3<\beta_1$ if $c > \hat{\gamma_2}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{result}[Equilibria and stability] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the cost boundaries $\gamma_1$ and $\gamma_2$ and the vertical transmission threshold $\hat v$ (\autoref{eq:cost_boundaries_v_threshold}) :

@@

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T_A \ge T_B$ and $c < \gamma_1$; or 
	if \emph{(ii)} $T_A < T_B$ and $v>\hat v$ and $c < \gamma_2$.
\item \emph{Fixation of defection}: 
    if \emph{(iii)} $T_A \ge T_B$ and $\gamma_2 < c$; or 
	if \emph{(iv)} $T_A < T_B$ and $\gamma_1 < c$.
\item \emph{Stable coexistence}: 
    if \emph{(v)} $T_A < T_B$ and $v<\hat{v}$ and $c < \gamma_1$; or 
    if \emph{(vi)} $T_A < T_B$ and $v>\hat{v}$ and $\gamma_2 < c < \gamma_1$.
\item \emph{Unstable coexistence}:
    if \emph{(vii)} $T_A > T_B$ and $\gamma_1 < c < \gamma_2$.
\end{enumerate}

@@

Much of the literature on evolution of cooperation focuses on conditions for  an initially rare cooperative phenotype to invade a population of defectors.
The following remarks on such a condition.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[Condition for cooperation to increase from rarity: cost of cooperation]
\label{remark:rarity}
If the initial frequency of cooperation is very close to zero, then its frequency will increase if the cost of cooperation is low enough,
\begin{equation} \label{eq:unequal_transmission_from_rarity_general_case}
c < \gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)} \;.
\end{equation} 
\end{remark}

This unites the conditions for fixation of cooperation and for stable coexistence, which both lead to instability of the fixation state associated with defection, $\hat{p}=0$.

Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation, $T_A$, increases the boundary ($\partial \gamma_1 / \partial T_A > 0$), facilitating the evolution of cooperation.
However, increasing the horizontal transmission of defection, $T_B$, can increase or decrease the cost boundary, but it increases the cost boundary when it is already above one ($c<1<\gamma_1$): $\partial\gamma_1/\partial T_B$ is positive when $T_A > \frac{1}{1+\alpha b v}$, which gives $\gamma_1>1/v$. 
Therefore, increasing $T_B$ decreases the cost boundary and limits the evolution of cooperation, but only if $T_A < \frac{1}{1+\alpha b v}$.

Increasing the vertical transmission rate, $v$, can either increase or decrease the boundary, depending on the horizontal transmission bias, $T_A-T_B$, because $\text{sign}(\partial \gamma_1 / \partial v) = -\text{sign}(T_A-T_B)$.
When $T_A<T_B$ we get $\partial \gamma_1 / \partial v >0$, and therefore as the vertical transmission rate increases, the cost boundary increases, making it easier for cooperation to evolve from rarity. 
In contrast, when $T_A > T_B$ we get $\partial \gamma_1 / \partial v <0$, and therefore as the vertical transmission rate increases, the cost boundary decreases, making it harder for cooperation to evolve from rarity.

In general, this condition cannot be formulated in the form of Hamilton's rule due to the horizontal transmission bias term, $T_A-T_B$.
When there is no horizontal transmission bias, $T_A=T_B$, the next remark applies.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[Unbiased horizontal transmission] 
\label{remark:equal_transmission}
If horizontal transmission is unbiased, $T=T_A=T_B$, then cooperation will take over the population from any initial frequency if the cost is low enough,
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{v \alpha T}{1-T} \;.
\end{equation}
\end{remark}
This condition is obtained by setting $T=T_A=T_B$ in inequality \ref{eq:unequal_transmission_from_rarity_general_case} and can be interpreted as a version of Hamilton's rule  ($c<b\cdot r$, inequality~\ref{eq:hamilton_rule}) or as a version of inequality~\ref{eq:fletcher2009}, where $v \alpha T/(1-T)$ can be regarded as the \emph{effective relatedness} or \emph{effective assortment}, respectively.
\autoref{fig:results}a illustrates this condition for $v=1$. Note that the right-hand side equals $\gamma_1$ when $T=T_A=T_B$.

%In Corollary~\autoref{remark:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
%A more general case can be stated for $0 < v \leq 1$.
In the next remarks we examine the effect of social association on the evolution of cooperation.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[No social association]
\label{remark:alpha0}
Without social association ($\alpha=0$), cooperation will increase from rarity if there is horizontal transmission bias for cooperation, $T_A>T_B$, and
\begin{equation}
\label{eq:vert_hori_alpha0}
c < \frac{T_A - T_B}{v(1-T_B)} \;.
\end{equation}
\end{remark}

\autoref{fig:result2}a illustrates this condition, which is obtained by setting $\alpha=0$ in inequality~\ref{eq:unequal_transmission_from_rarity_general_case}.
Importantly, the benefit of cooperation, $b$, does not affect the evolution of cooperation in the absence of social association, and the outcome is determined only by  cultural transmission.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[Perfect social association]
\label{remark:alpha1}
With perfect social association ($\alpha=1$), cooperation will increase from rarity if
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b v T_A + (T_A - T_B)}{v(1-T_B)} \;.
\end{equation}
\end{remark}
This condition is obtained by setting $\alpha=1$ in inequality~\ \ref{eq:unequal_transmission_from_rarity_general_case}.

@@

Given the previous two remarks for no association and perfect association ($\alpha=0$ and $\alpha=1$), it is interesting to examine the general effect of social association on the evolution of cooperation.
Consider the social association boundaries, $a_1$ and $a_2$,
\begin{equation} \label{eq:boundries_assortative_meeting_general_case}
  a_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  a_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{equation}
Then the following applies.
\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{remark}[Condition for cooperation to increase from rarity: social association]
\label{remark:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if $a_2 < \alpha$.
\end{remark}
Figures~\ref{fig:result2}c and \ref{fig:result3}cd illustrate this condition.
With horizontal transmission bias for cooperation, $T_A>T_B$, cooperation can fix from any initial frequency if $a_2<\alpha$ (green area in positive x-axis). 
With horizontal bias favoring defection, $T_A<T_B$, cooperation can fix from any frequency if social association is high, $a_1<\alpha$ (green area with $T_A<T_B$), and can also increase when rare and reach stable coexistence with defection if social association is intermediate, $a_2<\alpha$ (yellow area).
Without horizontal bias, $T_A=T_B$, fixation of cooperation occurs if social association is high enough, $\frac{c}{b \cdot v} \cdot \frac{1-T}{T} < \alpha$ (inequality~\ref{eq:equal_transmission}; in this case $a_1=a_2$).

Interestingly, because $\text{sign} (\partial a_2 / \partial v) = \text{sign} (T_A-T_B)$, the effect of the vertical transmission rate $v$ on this condition depends on the horizontal transmission bias. 
That is, with horizontal bias for cooperation, $T_A>T_B$, evolution of cooperation is facilitated by oblique transmission, whereas with horizontal bias for defection, $T_A<T_B$, evolution of cooperation is facilitated by vertical transmission.
This is demonstrated in \autoref{fig:result3}cd.

The next remarks further inspect the roles of vertical and oblique transmission in the evolution of cooperation.
\\
